Understanding Complex Patterns in Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models

### 2022-12-01 manuscript/manuscript.rmd
prepare for arxiv submission

---
title: Understanding Complex Patterns in Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models
author: 
  - name: "Christian Testa"
    institute: "1"
    email: "ctesta@hsph.harvard.edu"
    correspondence: true
institute: 
  - "1": "Department of Social and Behavioral Sciences, Harvard T.H. Chan School of Public Health"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
abstract: |
  I present three types of applications of generalized additive models (GAMs) to
  COVID-19 mortality rates in the US for the purpose of advancing methods to
  document inequities with respect to which communities suffered disproportionate
  COVID-19 mortality rates at specific times during the first three years of the 
  pandemic. First, GAMs can be used to describe the changing
  relationship between COVID-19 mortality and county-level covariates
  (sociodemographic, economic, and political metrics) over time. Second, GAMs can
  be used to perform spatiotemporal smoothing that pools information over time and
  space to address statistical instability due to small
  population counts or stochasticity can be summarized by a smooth, dynamic latent
  surface describing the mortality risk associated with geographic locations over
  time. Third, estimation of COVID-19 mortality associations with county-level covariates
  conditional on smooth spatiotemporal variation allows for more rigorous consideration
  of how socio-environmental contexts and policies may have impacted COVID-19 mortality.
  Each of these approaches
  provides a valuable perspective to documenting inequities in
  COVID-19 mortality by addressing the question of which populations have suffered
  the worst burden of COVID-19 mortality taking into account the nonlinear
  spatial, temporal, and social patterning of disease.
  \newline

  \noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
  Additive Model (GAM), Centers for Disease Control and Prevention (CDC), Index of Concentration at the Extremes (ICE)
bibliography: bibliography.bib
csl: apa.csl
link-citations: true
urlcolor: blue
linkcolor: black
header-includes: 
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
---

<!--
TODO:

Throughout/Overarching: 
  - Do I need to update all the models to separately control for age? Probably.
  - Do I want to have a supplementary materials section?
  - Talk about ecosocial theory (?) (Who benefits from harm done to whom?)

Abstract:
  - Add link to GitHub repository

Intro: 
  - add geographic background, like where the first cases were reported, 
  how COVID exploded in NYC, then the surges in the Midwest and South
  - add a primer on the relevance of spatiotemporal modeling

Methods: 
  - add intro to ICEraceinc (methods)
  - describe negative binomial models

Results:
  - add figure for temporal two-way interaction effects
  - present rate ratios (results)
  - present model comparisons (results, F-tests, AIC)

Discussion:
  - talk about how COVID has not been a win for health equity (discussion)
  - talk about limitations to being causal
  - talk about future directions

Conclusion:
  - summarize
-->

@@

Having passed over 1 million COVID-19 deaths in the United States earlier this
year [@donovan_us_2022], and facing uncertain prospects for the third COVID-19
winter looming even as new iterations on the COVID-19 vaccines become available,
it remains critical that inequities in COVID-19 outcomes are documented and
analyzed to reckon with the unjust and unfair burden of preventable illness.
Even though the first vaccines were granted emergency use authorization by the
U.S. Food and Drug Administration in 2020 [@mayo_clinic_history_2022], with the
first shots going in arms in December 2020, COVID-19 is still continuing to
cause hundreds of deaths a day in the US in the fall of 2022
[@noauthor_united_2022].  The new bivalent vaccines released at the end of
August 2022 contain mRNA sequences from both the original strain as well as
the recently emergent BA.4 and BA.5 lineages in an effort to make the nation's immunity more
up-to-date and robust against the myriad of phylogenetic directions the COVID-19
virus is evolving to explore [@office_of_the_commissioner_coronavirus_2022].
Despite the updated bivalent boosters representing a significant step forward in
prevention strategy, less than 4% of eligible Americans had taken the booster in
the first month after it became available [@bendix_less_2022; @lambert_most_2022]. 
As such, and with an enduring history of inequities in health care access 
in the US [@ortega_ending_2021; @blendon_inequities_2002; @bailey_how_2021;
@okonkwo_covid-19_2021; @rapp_statelevel_2022;
@chrisler_ageism_2016; @feldman_health_2021; @carpenter_health_2021;
@whitehead_outness_2016], it is clear that without further intervention not all
communities will be equally able to benefit from the updated vaccines and
inequities in COVID-19 illness and mortality may persist despite the
technological innovations in vaccine technology.

@@

```{r figure-county-trends, fig.cap = "Estimates of monthly COVID-19 mortality rates per 100,000 person-years by county organized by Census Division.  For each division, the median trendline and quantile ranges are shown weighted by county population size.", out.width = "6in", fig.align = 'center', fig.env = "figure*"}
knitr::include_graphics(here("analysis/02_plot_county_trends/county_trends_by_division.png"))
```

@@

Figure \@ref(fig:figure-county-trends) shows the monthly COVID-19 mortality
rates for the counties grouped within each of the nine U.S. Census Divisions
[@us_department_of_commerce_economics_and_statistics_administration_census_2000]. 
The figure summarizes each division's median mortality rates
weighted by county population size. Quantile ranges are included to illustrate
the range of mortality rates observed, again weighted by county population size.
Notably, the mortality associated
with the early surge of cases starting in NYC and spreading through New York,
New Jersey, and Massachusetts is visible in the Middle Atlantic and New England
division figures. The figure also illustrates how the first peak in the
mortality time-series for states in the Midwest (West North Central, East North
Central) occurred later, in late 2020 and going into early 2021.

In the US context, one of the key aspects to the geographic story of COVID-19's
spread and diffusion was the early surge of cases and epicenter in New York City
during March 2020 [@thompson_covid-19_2020] followed by subsequent waves of cases in the
South and Midwest [@shumaker_covid-19_2020; @scott_these_2020; @glenza_covid_2020]. As Park et al. 
stated summarized trends in the US from March 2020 to May 2021, "hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates" [-@park_covid-19_2021].

@@

Even since the beginning of the COVID-19 outbreak in the US, data reflected
sharp inequities in mortality rates.  During January 22nd to May 5th 2020,
county COVID-19 mortality rates were 4.94 (95% CI 4.78, 5.09) times higher in
counties in the highest quintile of percent People of Color (61%-100%) compared
to counties in the lowest quintile of percent People of Color (0%-17.2%)
[@chen_revealing_2021].  This was not wholly unanticipated: as COVID-19 was
beginning to take off in the US, some were already calling attention to the fact
that COVID-19 threatened to exacerbate existing disparities
[@kim_covid-19_2020]. Kim, Marrast, and Conigliaro noted three structural
barriers in COVID-19 prevention and care: 1) originally requiring New York
residents to have a doctor's prescription for a COVID-19 test reduced the
opportunity for healthcare for People of Color as they are less likely to have a
primary care provider; 2) drive-through testing made testing disadvantaged those
who rely on public transportation; 3) quarantining at home while waiting the
7-10 days originally required for COVID-19 test results to come back posed an
economic and social challenge that many in already financially difficult
situations may not have been able to take on [-@kim_covid-19_2020]. Others noted
yet more reasons why COVID-19 threatened to worsen an already inequitable
healthcare landscape in the US: those who reside in prisons and jails,
immigrants and undocumented people, people with disabilities, and people
experiencing homeless all face additional challenges in seeking and getting the
healthcare they deserve [@okonkwo_covid-19_2021]. Despite stay-at-home orders
that were prevalent in many states [@moreland_timing_2020], workers functioning in 
capacities essential to the functioning of critical infrastructure operations (later
termed "essential workers") were exposed to heightened risk of COVID-19 transmission 
[@national_conference_of_state_legislatures_covid-19_2021; @wei_risk_2022; @national_bureau_of_economic_research_measuring_2021; @the_lancet_plight_2020; @hanage_covid-19_2020].
It's not just a hypothetical that those with more structurally enfranchised
privileges have been more able to mitigate their risk of negative health
outcomes associated with COVID-19 — during February 1 to April 1, 2020, New York
City residents who lived in more affluent neighborhoods were more likely to have
left the city, while New York City residents from more Black and Hispanic
neighborhoods were more likely to continue working [@coven_disparities_2020].

@@

Generalized additive models (GAMs) improve upon generalized linear models by
allowing for the fitting of smooth functions that transform the $x$ variables.
This is a convenient means to include nonlinear relationships between the
outcome and predictor variables. Whereas a generalized linear model may look like 
$$ g(\mu_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} ...$$ 
a generalized additive model could look like 
$$ g(\mu_i) = \mathbf A_i \mathbf \theta + f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}, x_{i4}) + \dots$$
where $\mu_i \equiv \mathbb E(Y_i)$, $\mathbf A_i$ is a row of the model matrix 
for any strictly parametric model components, $\mathbf \theta$ is the corresponding
parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].

@@

knitr::include_graphics(here("analysis/10_supplementary_gam_figures/gam_demo_figure.png"))
```

@@

While a variety of varying kinds of spline bases exist (cubic splines, B- and
P-splines, thin-plate splines, etc.), our particular application setting
warrants the use of tensor-product smooths because we require anisotropic
penalization. That is to say, when we seek to create spatiotemporally smoothed
model estimates, it's inappropriate to assume that the amount of smoothing
across space should be the same as the amount of smoothing across time because
they're in fundamentally different units (space being measured in Cartesian
coordinates and time being measured in years, months, days, etc.). More details
about tensor-product smooths are available in Simon Wood's *Generalized Additive Models - An Introduction with R* [-@wood_generalized_2017].

@@

Besides age, which is a strong determinant of COVID-19 mortality risk, we have
included the following variables as covariates of interest:

  * Population Density per Square Mile
  * Median Household Income
  * Proportion in Poverty
  * the Index of Concentration at the Extremes for Racialized Economic Segregation 
  * Political Lean in the 2020 Election (1 = 100% Republican votes, -1 = 100% Democratic Votes)

@@

Applying the ICE approach to a specific context involves defining the axes of
privilege of interest.  In this case, data on racialized economic segregation
are used from the US Census American Community Survey on White households
earning more than $100,000 a year (the most privileged group) or households of
People of Color earning less than $25,000 a year (the least privileged group).
This variable is referred to throughout as the ICE for racialized economic
segregation, or `ICEraceinc` in the code.

Compared with the Gini
coefficient which is one of the most popular methods for summarizing area-level
rates of inequities, the ICE the major advantage that it is more suitable for
describing inequities at smaller area levels [@krieger_public_2016]. While the
Gini coefficient measures within-area dissimilarity (as in, for example, how
unequal wealth is distributed within a county), the ICE measure establishes
where on a spectrum a given county's population falls allowing for comparison
across counties. The Gini coefficient suffers from the fact that areas which are
made up of relatively homogeneous populations will appear as having low
within-area inequality (and therefore low Gini coefficient). Instead, the ICE
measures how much of the population is either privileged or not. The Gini
coefficient remains useful for reporting on the degree of inequity in larger
areal units (like countries, states, and regions), but at smaller areal units
(like counties, ZIP codes, census tracts) can be more difficult to interpret and
compare. Maps of ICE measure can elucidate what spatial social segregation and
polarization exist, and the ICE for racialized economic segregation has been
repeatedly and significantly associated with COVID-19 outcomes
[@hanage_covid-19_2020; @chen_revealing_2021; @krieger_relationship_2022;
@eichenbaum_health_2022; @brown_ecological_2021; @sonderlund_racialized_2022;
@saha_neighborhood-level_2020].

@@

The formula used puts a one-dimensional smoothing spline to represent a
nonlinear age-effect and a two-dimensional tensor-product smooth on the
interaction between the given covariate and the date. The `d=c(1,1)` argument
provides the instruction necessary to consider covariate and date as being on
separate scales and therefore to fit the tensor-product smooth with anisotropy —
that is, to allow for independent amounts of scaling in the dimensions of the
covariate and time. The offset used structures the regression to estimate rates
in units of person-time per 100,000 person-years. Since the death counts are
aggregated to the monthly level, the person-time in units of 100,000
person-years are calculated by taking each county's population size, dividing by
100,000, and dividing by 12 (for the 12 months in a year).

The above structure is used to estimate models for our different `covariate`
variables of interest: median income, percent in poverty, the ICE for racialized
economic segregation, political lean.  The model presenting median age treats
median age as the main covariate of interest including it as the `te(covariate,
date, d=c(1,1))`. Results of these models are summarized in Figure
\@ref(fig:one-variable-over-time) where the additional COVID-19 mortality
associated with each covariate is visualized.

@@

```{r one-variable-over-time, fig.cap = "Additional COVID-19 mortality associated with covariates over time. A) age and time, B) log (base 10) population density and time, C) median income and time, D) proportion in poverty and time, E) ICEraceinc and time, F) political lean and time. In panels B-F the effect of age is marginalized out using the median age in the US, 38.8 (ref:medianage).", fig.align = 'center', out.height = '75%'}
knitr::include_graphics(here("analysis/04_one_variable_over_time/panel_figure.png"))
```

@@

```{r two-variables-over-time, fig.cap="The changing interacted effect of the ICE for racialized economic segregation and age over time. A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022. An animated version is available online at \\url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md}"}
knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
```

@@

Figure \@ref(fig:crude-vs-smoothed) shows the difference between crude mortality
rates and spatially smoothed mortality rates in January 2022 to illustrate the
level of noise present in raw rates and how spatially smoothing synthesizes
local geographic patterns into interpretable trends that can be meaningfully
interpreted as local area mortality risk levels pooling information across
nearby county rates.

Figure \@ref(fig:spatiotemporal-panel-figure) shows the results from a
spatiotemporally smoothed model in select months to highlight how spatiotemporal
smoothing can yield results that intuitively synthesize trends in space and
time.

```{r crude-vs-smoothed, fig.cap = "Comparison of crude vs. smoothed mortality rates. A) Crude mortality rates per 100,000 person-years in January 2022. B) Smoothed mortality rates from a GAM applied to mortality rates from January 2022."}
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/side_by_side_crude_and_smoothed.png"))
```

```{r spatiotemporal-panel-figure, fig.cap = "Spatiotemporally smoothed COVID-19 mortality estimates from a GAM fit to data from March 2020-August 2022. Panels show months selected to highlight the changing spatial patterns of COVID-19 mortality risk over time. An animated version is available online at \\url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/09_spatiotemporal_models/animation/readme.md}"}
knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
```

@@

The final application of GAMs presented here is to estimate the effects of
covariates after adjusting for a spatiotemporally smoothed term as shown in
Application 2. This has the interpretation of asking what are the changes to
COVID-19 mortality rates associated with each covariate after taking into
account local geographic patterns over time. In particular, this is relevant for
understanding what county-level measures are associated with elevated COVID-19
mortality even after adjusting for where COVID-19 rates were locally elevated
or depressed due to variation in the timing of local introduction, transmission,
and diffusion.

```{r neighboring-graph-morans-i, fig.cap = "The spatial autocorrelation of COVID-19 rates in January 2022. Panel A shows the graph of which US counties are connected by neighboring relations. Panel B shows a scatterplot depicting each US counties' COVID-19 rate (x-axis) compared to the average of its neighbors' COVID-19 mortality rates (y-axis). A regression line shows the association between counties' COVID-19 rates and their neighboring counties' COVID-19 rates.  Dotted lines indicated the average for US county-level crude COVID-19 mortality rates (vertical) and for the average of each counties' neighboring counties' COVID-19 rates (horizontal)."}
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/county_network_and_morans_plot.png"))
```

```{r temporal-acf, fig.cap = "The median (black) and quantiles (shaded bands) of temporal autocorrelation of counties' COVID-19 mortality rates over time starting from April 2020.", out.width = '50%', fig.align='center'}
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/temporal_acf.png"))
```

To motivate the need for and application of spatiotemporal smoothing in
assessing the associated changes in COVID-19 mortality with county-level
covariates, Figure \@ref(fig:neighboring-graph-morans-i) presents the Moran's I
diagnostic plot for January 2022 which summarizes the amount of autocorrelation
between counties' COVID-19 mortality rates and a population-weighted average of
each counties' neighboring counties' COVID-19 mortality rates. The observed
autocorrelation indicates that there is an association between counties'
COVID-19 rates and their neighboring counties' COVID-19 rates, implying that
without taking this patterning into account, regression models that do not
explicitly model the effect of spatial autocorrelation may be biased due to
inappropriately assuming that county-level data are independent of one another. Second, 
while the Moran's I plot demonstrates the spatial-correlation present in the data
at a single time-point, the following Figure \@ref(fig:temporal-acf) shows the
tendency of county COVID-19 observations to be autocorrelated between subsequent
months and moreover there appears to be an oscillating trend at the 6-month and
12-month marks suggestive of effects due to the dynamics associated with
temporary infection induced immunity and seasonality.

>>>>

---
title: Understanding Complex Patterns in Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models
author: 
  - name: "Christian Testa"
    institute: "1"
    email: "ctesta@hsph.harvard.edu"
    correspondence: true
institute: 
  - "1": "Department of Social and Behavioral Sciences, Harvard T.H. Chan School of Public Health"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
    keep_tex: true
abstract: |
  I present three types of applications of generalized additive models (GAMs) to
  COVID-19 mortality rates in the US for the purpose of advancing methods to
  document inequities with respect to which communities suffered disproportionate
  COVID-19 mortality rates at specific times during the first three years of the 
  pandemic. First, GAMs can be used to describe the changing
  relationship between COVID-19 mortality and county-level covariates
  (sociodemographic, economic, and political metrics) over time. Second, GAMs can
  be used to perform spatiotemporal smoothing that pools information over time and
  space to address statistical instability due to small
  population counts or stochasticity resulting in a smooth, dynamic latent
  risk surface summarizing the mortality risk associated with geographic locations over
  time. Third, estimation of COVID-19 mortality associations with county-level covariates
  conditional on a smooth spatiotemporal risk surface allows for more rigorous consideration
  of how socio-environmental contexts and policies may have impacted COVID-19 mortality.
  Each of these approaches
  provides a valuable perspective to documenting inequities in
  COVID-19 mortality by addressing the question of which populations have suffered
  the worst burden of COVID-19 mortality taking into account the nonlinear
  spatial, temporal, and social patterning of disease.
  \newline

  \noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
  Additive Model (GAM), Centers for Disease Control and Prevention (CDC), Index of Concentration at the Extremes (ICE), Confidence Interval (CI)
bibliography: bibliography.bib
csl: apa.csl
link-citations: true
urlcolor: blue
linkcolor: black
header-includes: 
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
---

@@

Having passed over 1 million COVID-19 deaths in the United States in 2022
[@donovan_us_2022], and facing uncertain prospects for the third COVID-19 winter
looming even as new iterations on the COVID-19 vaccines become available, it
remains critical that inequities in COVID-19 outcomes are documented and
analyzed to reckon with the unjust and unfair burden of preventable illness.
Even though the first vaccines were granted emergency use authorization by the
U.S. Food and Drug Administration in 2020 [@mayo_clinic_history_2022], with the
first shots going in arms in December 2020, COVID-19 is still continuing to
cause hundreds of deaths a day in the US in the fall of 2022
[@noauthor_united_2022].  The new bivalent vaccines released at the end of
August 2022 contain mRNA sequences from both the original strain as well as the
recently emergent BA.4 and BA.5 lineages in an effort to make the nation's
immunity more up-to-date and robust against the myriad of phylogenetic
directions the COVID-19 virus is evolving to explore
[@office_of_the_commissioner_coronavirus_2022]. Despite the updated bivalent
boosters representing a significant step forward in prevention strategy, less
than 4% of eligible Americans had taken the booster in the first month after it
became available [@bendix_less_2022; @lambert_most_2022]. As such, and with an
enduring history of inequities in health care access in the US
[@ortega_ending_2021; @blendon_inequities_2002; @bailey_how_2021;
@okonkwo_covid-19_2021; @rapp_statelevel_2022;
@chrisler_ageism_2016; @feldman_health_2021; @carpenter_health_2021;
@whitehead_outness_2016], it is clear that without further intervention not all
communities will be equally able to benefit from the updated vaccines and
inequities in COVID-19 illness and mortality may persist despite the
technological innovations in vaccine technology.

@@

```{r figure-county-trends, fig.cap = "Estimates of monthly COVID-19 mortality rates per 100,000 person-years by county organized by Census Division.  For each division, the median trendline and quantile ranges are shown weighted by county population size.", out.width = "6in", fig.align = 'center', fig.env = "figure*"}
# knitr::include_graphics(here("analysis/02_plot_county_trends/county_trends_by_division.png"))
knitr::include_graphics(here("manuscript/figures/county_trends_by_division.png"))
```

@@

Figure \@ref(fig:figure-county-trends) shows the monthly COVID-19 mortality
rates for the counties grouped within each of the nine U.S. Census Divisions
[@us_department_of_commerce_economics_and_statistics_administration_census_2000]. 
The figure summarizes each division's median mortality rates
weighted by county population size. 
Notably, the mortality associated
with the early surge of cases starting in New York City and spreading through New York,
New Jersey, and Massachusetts is visible in the Middle Atlantic and New England
division figures. The figure also illustrates how the first peak in the
mortality time-series for states in the Midwest (West North Central, East North
Central) occurred later, in late 2020 and going into early 2021.

In the US context, one of the key aspects to the geographic story of COVID-19's
spread and diffusion was the early surge of cases and epicenter in New York City
during March 2020 [@thompson_covid-19_2020] followed by subsequent waves of cases in the
South and Midwest [@shumaker_covid-19_2020; @scott_these_2020; @glenza_covid_2020]. As Park et al.\ 
summarized the trends in the US from March 2020 to May 2021, "hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates" [-@park_covid-19_2021].

@@

Even since the beginning of the COVID-19 outbreak in the US, data reflected
sharp inequities in mortality rates.  During January 22nd to May 5th 2020,
county COVID-19 mortality rates were 4.94 (95% CI 4.78, 5.09) times higher in
counties in the highest quintile of percent People of Color (61%-100%) compared
to counties in the lowest quintile of percent People of Color (0%-17.2%)
[@chen_revealing_2021].  This was not wholly unanticipated: as COVID-19 was
beginning to take off in the US, some were already calling attention to the fact
that COVID-19 threatened to exacerbate existing disparities
[@kim_covid-19_2020]. Kim, Marrast, and Conigliaro noted at least three structural
barriers in COVID-19 prevention and care: 1) originally requiring
residents to have a doctor's prescription for a COVID-19 test reduced the
opportunity for healthcare for People of Color as they are less likely to have a
primary care provider; 2) drive-through testing made testing disadvantaged those
who rely on public transportation; and 3) quarantining at home while waiting the
7-10 days originally required for COVID-19 test results to come back posed an
economic and social challenge that many in already financially difficult
situations may not have been able to take on [-@kim_covid-19_2020]. Others noted
yet more reasons why COVID-19 threatened to worsen an already inequitable
healthcare landscape in the US: in particular, those who reside in prisons and jails,
immigrants and undocumented people, people with disabilities, and people
experiencing homeless all face additional challenges in seeking and getting the
healthcare they deserve [@okonkwo_covid-19_2021]. Even though stay-at-home orders
designed to mitigate spread that were prevalent in many states
[@moreland_timing_2020], workers functioning in capacities essential to the
functioning of critical infrastructure operations (later termed "essential
workers") were exposed to heightened risk of COVID-19 transmission
[@national_conference_of_state_legislatures_covid-19_2021; @wei_risk_2022;
@national_bureau_of_economic_research_measuring_2021; @the_lancet_plight_2020;
@hanage_covid-19_2020]. It's clear that those with more
structurally enfranchised privileges have been more able to mitigate their risk
of negative health outcomes associated with COVID-19 — during February 1 to
April 1, 2020, New York City residents who lived in more affluent neighborhoods
were more likely to have left the city, while New York City residents from more
Black and Hispanic
neighborhoods were more likely to continue working [@coven_disparities_2020].

@@

Generalized additive models (GAMs) improve upon generalized linear models by
allowing for the fitting of smooth functions that transform right-hand-side variables.
This is a convenient means to account for nonlinear relationships between the
outcome and predictor variables. Whereas a generalized linear model may look like 
$$ g(\mu_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} ...$$ 
a generalized additive model could look like 
$$ g(\mu_i) = \mathbf A_i \mathbf \theta + f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}, x_{i4}) + \dots$$
where the expected value of the outcome is given as $\mu_i \equiv \mathbb E(Y_i)$, 
$\mathbf A_i$ is a row of the model matrix 
for any strictly non-parametric model components, $\mathbf \theta$ is the corresponding
parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].

@@

# knitr::include_graphics(here("analysis/10_supplementary_gam_figures/gam_demo_figure.png"))
knitr::include_graphics(here("manuscript/figures/gam_demo_figure.png"))
```

@@

While a variety of spline-based approaches exist (cubic splines, B- and
P-splines, thin-plate splines, etc.), our particular application setting
warrants the use of tensor-product smooths because we require anisotropic
penalization. That is to say, when we seek to create spatiotemporally smoothed
model estimates, it's inappropriate to assume that the amount of smoothing
across space should be the same as the amount of smoothing across time because
they're in fundamentally different units (space being measured in Cartesian
coordinates and time being measured in years, months, days, etc.). More details
about tensor-product smooths are available in Simon Wood's *Generalized Additive Models - An Introduction with R* [-@wood_generalized_2017].

@@

The following variables have been included as covariates of interest:

  * Median Age
  * Population Density per Square Mile
  * Median Household Income
  * Proportion in Poverty
  * the Index of Concentration at the Extremes for Racialized Economic Segregation 
  * Political Lean in the 2020 Election (1 = 100% Republican votes, -1 = 100% Democratic Votes)

@@

Applying the ICE approach to a specific context involves defining the axes of
privilege of interest.  In this case, data on racialized economic segregation
are used from the US Census American Community Survey on white households
earning more than $100,000 a year (the most privileged group) or households of
People of Color earning less than $25,000 a year (the least privileged group).
This variable is referred to throughout as the ICE for racialized economic
segregation, or `ICEraceinc` in the code.

Compared with the Gini
coefficient which is one of the most popular methods for summarizing area-level
rates of inequities, the ICE has the advantage that it is suitable for
describing inequities at smaller area levels [@krieger_public_2016]. While the
Gini coefficient measures within-area dissimilarity (as in, for example, how
unequal wealth is distributed within a county), the ICE measure establishes
where on a spectrum a given county's population falls allowing for comparison
across counties. The Gini coefficient suffers from the fact that areas which are
made up of relatively homogeneous populations will appear as having low
within-area inequality (and therefore low Gini coefficient). Instead, the ICE
measures how much of the population is either privileged or not. The Gini
coefficient remains useful for reporting on the degree of inequity in larger
areal units (like countries, states, and regions), but at smaller areal units
(like counties, ZIP codes, census tracts) can be more difficult to interpret and
compare. Maps of ICE measure can elucidate what spatial social segregation and
polarization exist, and the ICE for racialized economic segregation has been
repeatedly and significantly associated with COVID-19 outcomes
[@hanage_covid-19_2020; @chen_revealing_2021; @krieger_relationship_2022;
@eichenbaum_health_2022; @brown_ecological_2021; @sonderlund_racialized_2022;
@saha_neighborhood-level_2020].

@@

The formula used puts a one-dimensional smoothing spline on median age to represent a
nonlinear age-effect and a two-dimensional tensor-product smooth on the
interaction between the given covariate and the date. The `d=c(1,1)` argument
provides the instruction necessary to consider covariate and date as being on
separate scales and therefore to fit the tensor-product smooth with anisotropy —
that is, to allow for independent amounts of scaling in the dimensions of the
covariate and time. The offset used structures the regression to estimate rates
in units of person-time per 100,000 person-years. Since the death counts are
aggregated to the monthly level, the person-time in units of 100,000
person-years are calculated by taking each county's population size, dividing by
100,000, and dividing by 12 (for the 12 months in a year).

The above structure is used to estimate models for our different `covariate`
variables of interest: median income, percent in poverty, the ICE for racialized
economic segregation, political lean.  The model presenting median age treats
median age as the main covariate of interest including it as the `te(covariate,
date, d=c(1,1))` and dropping the `s(median_age)` term which otherwise becomes redundant. 
Results of these models are summarized in Figure
\@ref(fig:one-variable-over-time) where the additional COVID-19 mortality
associated with each covariate is visualized.

@@

```{r one-variable-over-time, fig.cap = "Additional COVID-19 mortality associated with covariates over time. A) age and time, B) log (base 10) population density and time, C) median income and time, D) proportion in poverty and time, E) ICEraceinc and time, F) political lean and time. In panels B-F the effect of age is marginalized out using the median age in the US, 38.8 (ref:medianage).", fig.align = 'center', out.height = '75%'}
# knitr::include_graphics(here("analysis/04_one_variable_over_time/panel_figure.png"))
knitr::include_graphics(here("manuscript/figures/nonspatial_panel_figure.png"))
```

@@

```{r two-variables-over-time, fig.cap="The changing interacted effect of the ICE for racialized economic segregation and age over time. A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022. An animated version is available online at \\url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md}"}
# knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
knitr::include_graphics(here("manuscript/figures/ICEraceinc_age_panel_figure.png"))
```

@@

Figure \@ref(fig:crude-vs-smoothed) shows the difference between crude mortality
rates and spatially smoothed mortality rates in January 2022 to illustrate the
level of noise present in raw rates and how spatially smoothing synthesizes
local geographic patterns into trends that can be meaningfully
interpreted as local area mortality risk levels with information information pooled across
nearby county rates.

Figure \@ref(fig:spatiotemporal-panel-figure) shows the results from a
spatiotemporally smoothed model in select months to highlight how spatiotemporal
smoothing can yield results that synthesize trends in space and
time.

```{r crude-vs-smoothed, fig.cap = "Comparison of crude vs. smoothed mortality rates. A) Crude mortality rates per 100,000 person-years in January 2022. B) Smoothed mortality rates from a GAM applied to mortality rates from January 2022."}
# knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/side_by_side_crude_and_smoothed.png"))
knitr::include_graphics(here("manuscript/figures/side_by_side_crude_and_smoothed.png"))
```

```{r spatiotemporal-panel-figure, fig.cap = "Spatiotemporally smoothed COVID-19 mortality estimates from a GAM fit to data from March 2020-August 2022. Panels show months selected to highlight the changing spatial patterns of COVID-19 mortality risk over time. An animated version is available online at \\url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/09_spatiotemporal_models/animation/readme.md}"}
# knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
knitr::include_graphics(here("manuscript/figures/temporal_panel_figure.png"))
```

@@

The final application of GAMs presented here is to estimate the effects of
covariates after adjusting for a spatiotemporally smoothed latent risk surface. 
This has the interpretation of asking what are the changes to
COVID-19 mortality rates associated with each covariate after taking into
account regional geographic patterns in COVID-19 mortality over time. In
particular, this is relevant for understanding what county-level measures are
associated with elevated COVID-19 mortality even after adjusting for where
COVID-19 rates were locally elevated or depressed due to variation in the timing
of local introduction, transmission, and diffusion.

```{r neighboring-graph-morans-i, fig.cap = "The spatial autocorrelation of COVID-19 rates in January 2022. Panel A shows the graph of US counties connected according to which are neighbors. Panel B shows a scatterplot depicting each US counties' COVID-19 mortality rate (x-axis) compared to the average of its neighbors' COVID-19 mortality rates (y-axis). A regression line shows the association between counties' COVID-19 rates and their neighboring counties' COVID-19 mortality rates.  Dotted lines indicated the average for US county-level crude COVID-19 mortality rates (vertical) and for the average of each counties' neighboring counties' COVID-19 mortality rates (horizontal)."}
# knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/county_network_and_morans_plot.png"))
knitr::include_graphics(here("manuscript/figures/county_network_and_morans_plot.png"))
```

```{r temporal-acf, fig.cap = "The median (black) and quantiles (shaded bands) of temporal autocorrelation of counties' COVID-19 mortality rates over time during April 2020 — August 2022.", out.width = '50%', fig.align='center'}
# knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/temporal_acf.png"))
knitr::include_graphics(here("manuscript/figures/temporal_acf.png"))
```

To motivate the need for and application of spatiotemporal smoothing in
assessing the associated changes in COVID-19 mortality with county-level
covariates, Figure \@ref(fig:neighboring-graph-morans-i) presents the Moran's I
diagnostic plot for January 2022 which summarizes the amount of autocorrelation
between counties' COVID-19 mortality rates and the average of
each counties' neighboring counties' COVID-19 mortality rates. The observed
autocorrelation indicates that there is an association between counties'
COVID-19 rates and their neighboring counties' COVID-19 mortality rates, implying that
without taking this patterning into account, regression models that do not
explicitly model the effect of spatial autocorrelation may be biased due to
inappropriately assuming that county-level data are independent of one another. Second, 
while the Moran's I plot demonstrates the spatial-correlation present in the data
at a single time-point, the following Figure \@ref(fig:temporal-acf) shows the
tendency of county COVID-19 observations to be autocorrelated between subsequent
months.
